cardinalR: Generating interesting high-dimensional data structures

### 2025-11-08 paper-cardinalR.Rmd
Added fig.alt

```{r workflow, out.width = "70%", fig.pos="H", fig.align='center', fig.cap="Workflow for generating high-dimensional clustered data. The user specifies input parameters (number of points, clusters, cluster shapes, scaling, rotation, and optional background noise). Clusters are generated iteratively, transformed, optionally augmented with Gaussian noise dimensions, combined, and labeled, resulting in the final dataset.", fig.alt="Diagram showing the workflow for generating high-dimensional clustered data, from specifying input parameters to producing the final labeled dataset with optional noise dimensions."}

@@

```{r pyr-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `pyrrect` (a1-a3), `pyrtri` (b1-b3), `pyrstar` (c1-c3), and `pyrholes` (d1-d3) data. The `pyrrrect` structure forms a dense rectangular base tapering to a narrow tip, while `pytri` shows a more triangular spread with sharper edges. `Pyrstar` extends into multiple pointed branches radiating from a common core, and `pyrholes` reveals hollow or open regions within an otherwise compact shape. These projections illustrate a range of pyramid-like geometries that vary in density and structure.", fig.width=12, fig.height=16, fig.pos="H", out.width="80%", fig.align='center', fig.alt="Nine 2-D projections of four 4-D pyramid-like datasets: pyrrect, pyrtri, pyrstar, and pyrholes, showing their distinct geometries, branching, and hollow regions."}

@@

```{r sphere-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, `circle` (a1-a3), `curvycycle` (b1-b3), and, $3\\text{-}D$ `clusteredspheres` (c1-c3). The `circle` structure forms a smooth, closed loop, while `curvycycle` shows a wavy, continuous pattern forming a twisted ring. The `clusteredspheres` dataset displays multiple compact spherical groups that are clearly separated in higher dimensions but overlap slightly in some $2\\text{-}D$ projections, highlighting how projection can distort spatial relationships. These projections show how simple cyclic, wavy curvilinear, and clustered structures appear in $2\\text{-}D$, emphasizing the effects of projection on density, continuity, and separation", fig.width=12, fig.height=12, out.width="80%", fig.align='center', fig.pos="H", fig.alt="Nine 2-D projections of three datasets: circle, curvycycle, and clusteredspheres, showing closed loops, twisted rings, and compact spherical clusters, illustrating projection effects on structure, density, and separation."}

@@

```{r trefoil-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, `trefoil4d` (a1-a3) and $3\\text{-}D$ `trefoil3d` (b1-b3) data. The `trefoil4d` structure represents a higher-dimensional extension of the classic trefoil knot, revealing complex twisting and looping patterns that remain continuous across projections. In contrast, the `trefoil3d` dataset maintains a simpler, more compact knot-like form, showing how dimensional extension adds curvature and separation in the embedded space. These projections illustrate a range of looping structures in high-dimensions.", fig.width=12, fig.height=8, fig.pos="H", out.width="80%", fig.align='center', fig.alt="Six 2-D projections of trefoil knot datasets, showing complex 4-D trefoil loops and simpler 3-D knot structures, illustrating curvature, twisting, and separation across dimensions."}

@@

```{r triginometric-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `curvycylinder` (a1-a3), `sphericalspiral` (b1-b3), `conicspiral` (c1-c3), and `nonlinear` (d1-d3) data. The `curvycylinder` shows a cylindrical manifold with a nonlinear twist along its height, producing smooth, continuous curvature. The `sphericalspiral` forms a spiral path on a spherical surface, combining circular and vertical motion in a helical form. The `conicspiral` spreads radially while ascending, forming a conical helix with twisting variations in a non-radial dimension. The `nonlinear` dataset exhibits a warped $2\\text{-}D$ surface with sharp oscillations and smooth waves, reflecting complex nonlinear interactions. Each shows variations in curvature, density, and continuity.", fig.width=12, fig.height=16, out.width="70%", fig.align='center', fig.pos="H", fig.alt="Twelve 2-D projections of four 4-D datasets: curvycylinder, sphericalspiral, conicspiral, and nonlinear, showing cylindrical twists, spiral and conical helices, and warped 2-D surfaces, illustrating curvature, density, and continuity."}

>>>>

```{r workflow, out.width = "70%", fig.pos="H", fig.align='center', fig.cap="Workflow for generating high-dimensional clustered data. The user specifies input parameters (number of points, clusters, cluster shapes, scaling, rotation, and optional background noise). Clusters are generated iteratively, transformed, optionally augmented with Gaussian noise dimensions, combined, and labeled, resulting in the final dataset.", fig.alt="Workflow for generating high-dimensional clustered data, from specifying input parameters to producing the final labeled dataset with optional noise dimensions."}

@@

```{r pyr-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `pyrrect` (a1-a3), `pyrtri` (b1-b3), `pyrstar` (c1-c3), and `pyrholes` (d1-d3) data. The `pyrrrect` structure forms a dense rectangular base tapering to a narrow tip, while `pytri` shows a more triangular spread with sharper edges. `Pyrstar` extends into multiple pointed branches radiating from a common core, and `pyrholes` reveals hollow or open regions within an otherwise compact shape. These projections illustrate a range of pyramid-like geometries that vary in density and structure.", fig.width=12, fig.height=16, fig.pos="H", out.width="80%", fig.align='center', fig.alt="Three 2-D projections from each of four 4-D pyramid-like datasets: pyrrect, pyrtri, pyrstar, and pyrholes, showing their distinct geometries, branching, and hollow regions."}

@@

```{r sphere-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, `circle` (a1-a3), `curvycycle` (b1-b3), and, $3\\text{-}D$ `clusteredspheres` (c1-c3). The `circle` structure forms a smooth, closed loop, while `curvycycle` shows a wavy, continuous pattern forming a twisted ring. The `clusteredspheres` dataset displays multiple compact spherical groups that are clearly separated in higher dimensions but overlap slightly in some $2\\text{-}D$ projections, highlighting how projection can distort spatial relationships. These projections show how simple cyclic, wavy curvilinear, and clustered structures appear in $2\\text{-}D$, emphasizing the effects of projection on density, continuity, and separation", fig.width=12, fig.height=12, out.width="80%", fig.align='center', fig.pos="H", fig.alt="Three 2-D projections from each of three datasets: circle, curvycycle, and clusteredspheres, showing closed loops, twisted rings, and compact spherical clusters, illustrating projection effects on structure, density, and separation."}

@@

```{r trefoil-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, `trefoil4d` (a1-a3) and $3\\text{-}D$ `trefoil3d` (b1-b3) data. The `trefoil4d` structure represents a higher-dimensional extension of the classic trefoil knot, revealing complex twisting and looping patterns that remain continuous across projections. In contrast, the `trefoil3d` dataset maintains a simpler, more compact knot-like form, showing how dimensional extension adds curvature and separation in the embedded space. These projections illustrate a range of looping structures in high-dimensions.", fig.width=12, fig.height=8, fig.pos="H", out.width="80%", fig.align='center', fig.alt="Three 2-D projections from each of two trefoil knot datasets, showing complex 4-D trefoil loops and simpler 3-D knot structures, illustrating curvature, twisting, and separation across dimensions."}

@@

```{r triginometric-proj, eval=knitr::is_latex_output(), fig.cap="Three $2\\text{-}D$ projections from $4\\text{-}D$, for the `curvycylinder` (a1-a3), `sphericalspiral` (b1-b3), `conicspiral` (c1-c3), and `nonlinear` (d1-d3) data. The `curvycylinder` shows a cylindrical manifold with a nonlinear twist along its height, producing smooth, continuous curvature. The `sphericalspiral` forms a spiral path on a spherical surface, combining circular and vertical motion in a helical form. The `conicspiral` spreads radially while ascending, forming a conical helix with twisting variations in a non-radial dimension. The `nonlinear` dataset exhibits a warped $2\\text{-}D$ surface with sharp oscillations and smooth waves, reflecting complex nonlinear interactions. Each shows variations in curvature, density, and continuity.", fig.width=12, fig.height=16, out.width="70%", fig.align='center', fig.pos="H", fig.alt="Three 2-D projections from each of four 4-D datasets: curvycylinder, sphericalspiral, conicspiral, and nonlinear, showing cylindrical twists, spiral and conical helices, and warped 2-D surfaces, illustrating curvature, density, and continuity."}

### 2025-11-08 paper-cardinalR.tex
Added fig.alt

{\centering \includegraphics[width=0.7\linewidth,alt={Diagram showing the workflow for generating high-dimensional clustered data, from specifying input parameters to producing the final labeled dataset with optional noise dimensions.}]{figures/cardinalR_workflow}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Nine 2-D projections of four 4-D pyramid-like datasets: pyrrect, pyrtri, pyrstar, and pyrholes, showing their distinct geometries, branching, and hollow regions.}]{paper-cardinalR_files/figure-latex/pyr-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Nine 2-D projections of three datasets: circle, curvycycle, and clusteredspheres, showing closed loops, twisted rings, and compact spherical clusters, illustrating projection effects on structure, density, and separation.}]{paper-cardinalR_files/figure-latex/sphere-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Six 2-D projections of trefoil knot datasets, showing complex 4-D trefoil loops and simpler 3-D knot structures, illustrating curvature, twisting, and separation across dimensions.}]{paper-cardinalR_files/figure-latex/trefoil-proj-1}

@@

{\centering \includegraphics[width=0.7\linewidth,alt={Twelve 2-D projections of four 4-D datasets: curvycylinder, sphericalspiral, conicspiral, and nonlinear, showing cylindrical twists, spiral and conical helices, and warped 2-D surfaces, illustrating curvature, density, and continuity.}]{paper-cardinalR_files/figure-latex/triginometric-proj-1}

>>>>

{\centering \includegraphics[width=0.7\linewidth,alt={Workflow for generating high-dimensional clustered data, from specifying input parameters to producing the final labeled dataset with optional noise dimensions.}]{figures/cardinalR_workflow}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Three 2-D projections from each of four 4-D pyramid-like datasets: pyrrect, pyrtri, pyrstar, and pyrholes, showing their distinct geometries, branching, and hollow regions.}]{paper-cardinalR_files/figure-latex/pyr-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Three 2-D projections from each of three datasets: circle, curvycycle, and clusteredspheres, showing closed loops, twisted rings, and compact spherical clusters, illustrating projection effects on structure, density, and separation.}]{paper-cardinalR_files/figure-latex/sphere-proj-1}

@@

{\centering \includegraphics[width=0.8\linewidth,alt={Three 2-D projections from each of two trefoil knot datasets, showing complex 4-D trefoil loops and simpler 3-D knot structures, illustrating curvature, twisting, and separation across dimensions.}]{paper-cardinalR_files/figure-latex/trefoil-proj-1}

@@

{\centering \includegraphics[width=0.7\linewidth,alt={Three 2-D projections from each of four 4-D datasets: curvycylinder, sphericalspiral, conicspiral, and nonlinear, showing cylindrical twists, spiral and conical helices, and warped 2-D surfaces, illustrating curvature, density, and continuity.}]{paper-cardinalR_files/figure-latex/triginometric-proj-1}
